---
title: "Analyse ABC Results"
format: html
editor: visual
---

```{r, libraries}
#| echo: false
#| warning: false
library(ggplot2)  
library(knitr) 
library(magrittr) 
library(dplyr)
library(tidyr)
library(msm) # For truncated normal distribution
library(MASS) # To estimate distribution values
library(foreach)
library(doParallel)
```

## Data load

We have pre-run the data on HPC, as detailed in `RunABC.qmd`. We load this, and the experimental data for comparison.

```{r dataload}
load("data/ABCSummaryStatistic-1e3samples.RData")
obs.long <- read.csv("data/expdata.csv")
obs.wide <- rename(obs.long,"t"="week") %>%
    pivot_wider(names_from="compartment", values_from="N", values_fill = 0)
```

## Summary statistic

We have pre-calculated our summary statistic for each compartment: sum of the square error

$$
S_i=\sum_j \sum_k \left(x_{ik,obs}(t_j)-x_{i,sim}(t_j|\theta)\right)^2
$$

for compartment $x_i$ in observed data $X_{obs}$ (for each experiment $k$, 4 total) and simulated data $X_{sim}$ for proposed parameter set $\theta$ at time $t_j$ over all observation time points ($T = \{1,2,4,6,8,10,15\}$ weeks).

### Error per compartment

Let's have a look at the error in each compartment to see what we can predict well and what we can't predict well:

```{r error-distr}
pivot_longer(Si,-iteration) %>%
  ggplot(aes(x=value)) +
  geom_histogram() +
  facet_wrap(~name, scales="free") +
  scale_y_log10()
```

## Acceptance criteria

### Threshold all compartments

If we apply some threshold to all compartments we get:

```{r}
#| echo: false
#| warning: false
acceptedP <- list()
acceptanceRate <- list()
lognormalFit <- list()
state_cols = names(Si)[names(Si) != "iteration"]
for (threshold in seq(1400,2000,200)) {
  acceptance <- apply(Si, 1, function(si) 
                                  all( si[state_cols] < threshold) )
  acceptedP[[as.character(threshold)]] <- parametersets[acceptance,]
  acceptanceRate[[as.character(threshold)]] = 100*mean(acceptance)
  if (sum(acceptance) > 0 ) {
    lognormalFit[[as.character(threshold)]] = apply(parametersets[acceptance,],2,fitdistr,densfun="lognormal")
  }
}
acceptedP <- bind_rows(acceptedP, .id="threshold")
# Print acceptance rate
kable(t(unlist(acceptanceRate)), caption="Percent accepted for different threshold values. ")

# Determine the priors for density overlay
prior <- bind_rows(parametersets) %>%
  pivot_longer(cols=everything()) %>%
  mutate(threshold = "Prior")
# Plot
pivot_longer(acceptedP, cols=-threshold) %>%
  ggplot(aes(x=value)) +
  geom_histogram(aes(fill=threshold, colour=threshold), alpha=0.2, bins=50) + 
  facet_grid(threshold~name, scales="free") +
  geom_histogram(data=prior, bins=50)

# Print the fit estimates
lapply(lognormalFit, function(f) sapply(f,"[[","estimate")) %>%
  lapply(kable)
```

Let's also look at a pair plot for this acceptance method (threshold of 5, taking the first 1000 samples only):

```{r}
#| echo: false
#| warning: false
library(GGally)
x <- acceptedP[acceptedP$threshold == 1600,]
ggpairs(x[1:1e3,-1],
        diag=list(continuous=wrap("barDiag",colour="black",fill="lightgrey")),
        lower = list(continuous=wrap("points",alpha=0.3,size=0.5)))
```

### Threshold average across compartments

If we take the mean difference across all compartments and threshold on this we get:

```{r mean-val}
#| echo: false
#| warning: false
acceptedP <- list()
acceptanceRate <- list()
state_cols = names(Si)[names(Si) != "iteration"]
for (threshold in seq(400,500,20)) {
  acceptance <- apply(Si, 1, function(si) all(mean(si[state_cols]) < threshold))
  acceptedP[[as.character(threshold)]] <- parametersets[acceptance,]
  acceptanceRate[[as.character(threshold)]] = 100*mean(acceptance)
}
acceptedP <- bind_rows(acceptedP, .id="threshold")
# Print acceptance rate
kable(t(unlist(acceptanceRate)), caption="Percent accepted for different threshold values. ")

# Determine the priors for density overlay
prior <- bind_rows(parametersets) %>%
  pivot_longer(cols=everything()) %>%
  mutate(threshold = "Prior")
# Plot
pivot_longer(acceptedP, cols=-threshold) %>%
  ggplot(aes(x=value)) +
  geom_histogram(aes(fill=threshold, colour=threshold), alpha=0.2, bins=50) + 
  facet_grid(threshold~name, scales="free") +
  geom_histogram(data=prior, bins=50)
```

### Threshold specific compartments

Finally we can look at just I2/IV2:

```{r}
#| echo: false
#| warning: false
acceptedP <- list()
acceptanceRate <- list()
for (threshold in seq(400,500,20)) {
  acceptance <- apply(Si, 1, function(si) (si[["I2"]] <= threshold & si[["IV2"]] <= threshold))
  acceptedP[[as.character(threshold)]] <- parametersets[acceptance,]
  acceptanceRate[[as.character(threshold)]] = 100*mean(acceptance)
}
acceptedP <- bind_rows(acceptedP, .id="threshold")
# Print acceptance rate
kable(t(unlist(acceptanceRate)), caption="Percent accepted for different threshold values. ")

# Determine the priors for density overlay
prior <- bind_rows(parametersets) %>%
  pivot_longer(cols=everything()) %>%
  mutate(threshold = "Prior")
# Plot
pivot_longer(acceptedP, cols=-threshold) %>%
  ggplot(aes(x=value)) +
  geom_histogram(aes(fill=threshold, colour=threshold), alpha=0.2, bins=50) + 
  facet_grid(threshold~name, scales="free") +
  geom_histogram(data=prior, bins=50)
```
