---
title: "Modelling results for manuscript: 'Hotspot shelters enable frogs to survive chytridiomycosis and stimulate resistance'"
author: "Claire Miller, Jennifer Flegg, Patricia Campbell"
date: "Feb. 2024"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
# Data and model filepaths
expdata_filename <- "expdata.csv"
source("chytrid_models_and_residuals.R")
source("plotting_functions.R")

# Libraries
knitr::opts_chunk$set(echo = F, warning = F)
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)
library(RColorBrewer)
library(minpack.lm) # For the fitting
library(deSolve)
library(tibble)

# Change the ggplot theme
theme_set(theme_bw(base_size = 12))
options(ggplot2.discrete.colour = brewer.pal(7, "Dark2"), ggplot2.discrete.fill = brewer.pal(7, "Dark2"))
```


This RMarkdown report contains all the modelling results for the manuscript: Waddle et al. (2024) "Hotspot shelters enable frogs to survive chytridiomycosis and stimulate resistance".
The report is divided into the following sections, associated tables and figures from the manuscript are indicated in the parentheses:

1. [Model equations](#model-equations)
2. [Experimental data](#exp-data)
3. [Parameter fitting](#model-fitting) (Table 1)
4. [Fit comparison to data](#model-validation) (Figure S6)
5. [Sensitivity of fitting](#sensitivity) (Figure S7)
6. [Truncated normal fit](#truncated-normal) (Figure S8)
7. [Stochastic trajectories](#stochastic-trajectories) (Figure S9)

## Model equations <a name="model-equations"></a>

The model equations are: $$\begin{aligned}
  \frac{dS}{dt}   &= -\frac{\lambda S}{(N-1)} - \mu S \\
  \frac{dI_1}{dt}  &= \frac{\lambda S}{(N-1)} - (\gamma_1+\mu) I_1 \\
  \frac{dI_2}{dt}  &= \gamma_1 I_1 - (\gamma_2+\mu) I_2 \\
  \frac{dI_3}{dt}  &= \gamma_2 I_2 - (\omega+\mu) I_3 \\
  \frac{dS_V}{dt}  &= -\alpha \frac{\lambda S_V}{(N-1)} + \omega I_{V3} + \omega I_3 - \mu S_V \\
  \frac{dI_{V1}}{dt} &= \alpha \frac{\lambda S_V}{(N-1)} - (\gamma_1+\mu) I_{V1} \\
  \frac{dI_{V2}}{dt} &= \gamma_1 I_{V1} - (\gamma_2+\mu) I_{V2} \\
  \frac{dI_{V3}}{dt} &= \gamma_2 I_{V2} - (\omega+\mu) I_{V3} \\
  \frac{dN}{dt} &= -\mu N
\end{aligned}$$
where
$$
\lambda = \beta(m_{int}I_1+h_{int}I_2+I_3 + m_{int,V}I_{V1}+h_{int,V}I_{V2}+l_{int,V}I_{V3})
$$

for: $S$ the number of susceptible frogs; $I_1$, $I_2$, and $I_3$ are the number of infected frogs in the first, second, and third stages of infection, respectively; $S_V$ the number of vaccinated susceptible frogs; $I_{V1}$, $I_{V2}$, and $I_{V3}$ the number of vaccinated frogs in the first, second, and third stages of infection, respectively, and the total number of frogs $N$.

## Experimental data <a name="exp-data"></a>

```{r read-exp-data}
# Read in the two experimental datasets
# per-compartment timeseries, and per-frog data
# We expect the per-compartment data to have the following columns:
#   - Mesocosm: the mesocosm ID
#   - shaded: true/false whether the mesocosm was shaded or unshaded
#   - week: the observation week
#   - compartment: the compartment label
#   - N: the number of frogs cross all compartments
expdata <- read.csv(expdata_filename,check.names=F) %>%
  mutate(Mesocosm=as.factor(Mesocosm)) %>%
  pivot_wider(names_from=compartment,values_from=N) %>%
  mutate(across(-c(Mesocosm,shaded,week),~ifelse(is.na(.x),0,.x))) %>%
  rowwise() %>%
  mutate(N = sum(c_across(-c(Mesocosm,shaded,week)))) %>%
  ungroup()
```

The data is from from `r expdata_filename`, which was processed from the raw data using the `process_frog_compartments.R` script.
We assume at T=0 there were 5 frogs in each of the following categories:

- infect.+vacc. (IV1)
- susc.+vacc. (SV)
- infect. (I1)
- susc. (S)

and all other compartments had 0 frogs.

The following parameters for the model were calculated directly from laboratory data (Table S3 in manuscript):
```{r}
# Print parameters, except beta, alpha, mu, and omega
fixed_parms <- parameters[!(names(parameters) %in% c('mu','beta','alpha','omega'))]
kable(t(fixed_parms), caption = "Fixed parameters")
```


## Parameter fitting <a name="model-fitting"></a>

Parameter fitting was performed using the `minpack.lm` package. 
Full details on the fitting methods are given in the manuscript.

```{r fit-params}
# Original guesses are given in the model script
unknown_params = list(
  beta_sh = parameters[["beta"]],
  beta_un = parameters[["beta"]],
  alpha = parameters[["alpha"]], 
  omega = parameters[["omega"]],
  mu = parameters[["mu"]]
)

# Fit model
fit=nls.lm(par=unknown_params,lower=c(0,0,0,0,0),
           fn=calc_residuals,expdata=expdata)
summfit <- summary(fit)
fit_coefficients <- summfit$coefficients
```

The fitted parameters Estimate and 95% CI (Table 1 in the article) are as follows:
```{r table-1}
full_join(
  rownames_to_column(data.frame(fit_coefficients), var="Parameter"),
  rownames_to_column(data.frame(confint(fit),check.names=F), var="Parameter"),
  by="Parameter"
) %>%
  mutate(`95% CI` = paste0("(",round(`2.5 %`,digits=3),", ",
                           round(`97.5 %`,digits=3),")")) %>%
  select(Parameter,Estimate,`95% CI`) %>%
  kable(digits=3)
```


## Fit comparison to data <a name="model-validation"></a>

The following plots show the deterministic prediction, given the estimates above, against the experimental data (Figure S6 in manuscript).

```{r shaded-comparison}
# Shaded plots
parms_shaded = c(fixed_parms,
  fit_coefficients[rownames(fit_coefficients) != "beta_un","Estimate"])
names(parms_shaded) = gsub("beta_sh","beta",names(parms_shaded))
plot_compartments(list(parms_shaded), filter(expdata,shaded)) +
  labs(title="Shaded")
```

```{r unshaded-comparison}
# Unshaded plots
parms_unshaded = c(fixed_parms,
  fit_coefficients[rownames(fit_coefficients) != "beta_sh","Estimate"])
names(parms_unshaded) = gsub("beta_un","beta",names(parms_shaded))
plot_compartments(list(parms_unshaded), filter(expdata,!shaded)) +
  labs(title="Unshaded")
```


## Sensitivity analysis <a name="sensitivity"></a>

We explore the sensitivity of the parameter fits to the data by leaving each mesocosm out in turn and refitting the model.
The plot shows the estimates and 95% CI for each mesocosm removed (Figure S7 in manuscript). 

```{r sensitivity-analysis}
# Run for each mesocosm removed, 0: no mesocosms removed
res.sa <- c() 
for (mesocosm_i in c("None",levels(expdata$Mesocosm))) {
  # Fit model
  data_i <- filter(expdata, Mesocosm != mesocosm_i)
  fit_i=nls.lm(par=unknown_params,
             lower=c(0,0,0,0,0),
             fn=calc_residuals,
             expdata=data_i)
  fit_coefficients_i <- summary(fit_i)$coefficients
  
  # Extract the estimate and 95% CI
  parms_i <- full_join(
    rownames_to_column(data.frame(fit_coefficients_i), var="Parameter"),
    rownames_to_column(data.frame(confint(fit_i),check.names=F), var="Parameter"),
    by="Parameter"
    ) %>%
    mutate(MesocosmRm = mesocosm_i)

  # Append to results data frame
  res.sa <- rbind(res.sa,
                  select(parms_i, Parameter,Estimate,`2.5 %`,`97.5 %`,MesocosmRm))
}
```

```{r sa-plot}
# Process data for plotting
res.sa$ParameterMath = gsub("(beta)_([a-z][a-z]).*","\\1[\\2]",res.sa$Parameter)
res.sa$MesocosmRm = factor(res.sa$MesocosmRm, levels=c("None",levels(expdata$Mesocosm)))

ggplot(res.sa,
       aes(color=ParameterMath, x=MesocosmRm, y=Estimate, ymin=`2.5 %`, ymax=`97.5 %`)) +
  geom_errorbar(position=position_dodge(width=0.5), size =  1) +
  geom_point(position=position_dodge(width=0.5), size=3) +
  labs(x="Mesocosm removed", y="Estimate") +
  scale_color_discrete(name="Parameter", labels= function(x) parse(text = x))

```

## Truncated normal fit <a name="truncated-normal"></a>
We compare the fits using a truncated normal distribution to the data rather than a least squares (Figure S8 in manuscrpt).

```{r trunc-normal-fit}
# Run the fit
fit <- optim(par = unknown_params, fn = calc_residuals_obsmodel, 
             expdata = expdata, method = "SANN",
             lower = rep(0.001,length(unknown_params)),
             upper = c(rep(0.5,length(unknown_params)-1), 0.05))

# Print the coefficients compared to the least squares fit
data.frame(`Truncated Normal` = fit$par,
            `Least squares (above)` = fit_coefficients[,"Estimate"],
            check.names=F) %>%
  kable(digits=3)
```

```{r trunc-normal-shaded}
# Truncated normal fit shaded
parms_shaded_tn = c(fixed_parms,
  fit$par[names(fit$par) != "beta_sh"])
names(parms_shaded_tn) = gsub("beta_un","beta",names(parms_shaded))
plot_compartments(list(parms_shaded, parms_shaded_tn), 
                  filter(expdata,shaded), 
                  c("Least squares", "Truncated normal")) +
  labs(title="Shaded")
```

## Stochastic trajectories <a name="stochastic-trajectories"></a>

Example trajectories of the number of frogs are simulated using the estimated parameters and a Gillespie algorithm from the `GillespieSSA` library (Figure S9 in manuscrpt).

```{r gillespie-trajectories}
set.seed(141)
plot_gillespie(parms_shaded) + labs(title="Shaded")
plot_gillespie(parms_unshaded) + labs(title="Unshaded")
```

