---
title: "Run an ABC for the Chytrid Model"
format: html
editor: visual
---

```{r, libraries}
#| echo: false
#| warning: false
library(ggplot2)  # For plotting
library(knitr)    # For table output
library(magrittr) 
library(GillespieSSA) # To run gillespie trajectories
library(dplyr)
library(tidyr)
library(msm) # For truncated normal distribution
# Parallel running
library(foreach)
library(doParallel) 
set.seed(1) # Set the seed (for gillespie)
```

This is a useful review article: *Statistical inference for stochastic simulation models- theory and application* (Hartig et al. 2011, Ecology Letters), and also the blog posts Michael recommended from the paper's author Florian Hartig (a [simple ABC-MCMC](https://theoreticalecology.wordpress.com/2012/07/15/a-simple-approximate-bayesian-computation-mcmc-abc-mcmc-in-r/) and a [Metrolopolis-Hastings MCMC](https://theoreticalecology.wordpress.com/2010/09/17/metropolis-hastings-mcmc-in-r/)).

## Data

First process the observational data (and plot just to remind ourselves). For now we let's just look at the shaded environment.

```{r, dataload}
#| echo: false
# Read in data
obs.long <- read.csv("data/expdata.csv")
# Change the time column from week to t to match function nomenclature
obs <- rename(obs.long,"t"="week") %>%
# We want to match the wide format with one time column 't' and one column per compartment
    pivot_wider(names_from="compartment", values_from="N", values_fill = 0) %>%
    select(-Mesocosm)
# Plot
obs.long$Mesocosm = as.factor(obs.long$Mesocosm)
obs.long$compartment = factor(obs.long$compartment,levels=c("S","I1","I2","I3","SV","IV1","IV2","IV3"))
ggplot(obs.long,aes(x=week, y=N, group=Mesocosm, colour=shaded)) +
  geom_line() + geom_point(alpha = 0.6) +
  facet_grid(shaded~compartment) +
  labs(y="No. frogs")
```

## Functions required to run ABC

### Run Model

We use the Gillespie algorithm to run the model, using the GillespieSSA library.

```{r, modelrundef}
# Source the ODE system from the R script
source("chytrid_models_and_residuals.R")
# Get the gillespie system using the function from this script
gillespie_system <- build_system_gillespie()

# A function that solves the ODE model for a certain set of params
# @param params the parameter guess we want to run
# @param Tvec a vector of times we want to get the output for
# @return data.frame of the compartment counts at requested times
runModel <- function(params.fit, Tvec)
{
  # Initial conditions
  x0 <- c( S=5,  I1=5,  I2=0,  I3=0, 
           SV=5, IV1=5, IV2=0, IV3=0)
  # Combine fitting parameters with other model parameters 
  # (These are defined in chytrid_models_and_residuals.R)
  parameters[names(params.fit)] = params.fit
  # Run a trajectory
  trajectory <- ssa(x0 = x0,
      a = gillespie_system[["a"]],
      nu = gillespie_system[["nu"]],
      parms = parameters, 
      tf = max(Tvec),
      method = ssa.d(),verbose = FALSE, consoleInterval = 1)
  # Extract the data at the correct time points
  Xi <- lapply(setNames(Tvec,Tvec),function(ti, data) {
    # Determine the appropriate row in the solution (max is in case ti=0)
    rowi = max(which(data[,"t"] >= ti)[1]-1,1)
    # If it returns NA it will be the last time point
    rowi = ifelse(is.na(rowi), nrow(data), rowi)
    return(data[rowi,])
  }, data=trajectory$data)
  Xi <- bind_rows(Xi, .id="t") # override the t column with obs time
  return(Xi)
}

# A function that solves the ODE model over 2 parameter sets-one per environment (shaded: beta_sh and unshaded: beta_un)
# @param params the parameter guesses we want to run
# @param Tvec a vector of times we want to get the output for
# @return data.frame of the compartment counts at requested times
runModelBothEnv <- function(params.fit, Tvec)
{
  # Environment independent parameters
  params.env <- params.fit[grep("beta",names(params.fit),invert=T)]
  # Run shaded model
  params.shaded = c(params.env, beta=params.fit[["beta_sh"]])
  Xsim_sh <- runModel(params.shaded, Tvec)
  # Run unshaded model
  params.unshaded = c(params.env, beta=params.fit[["beta_un"]])
  Xsim_un <- runModel(params.unshaded, Tvec)
  # Return both as dataframe with boolean column for shaded or not
  Xsim_sh$shaded = T
  Xsim_un$shaded = F
  return(rbind(Xsim_sh, Xsim_un))
}
```

### Summary statistic

Now we need to choose the summary statistic(s) we want to use to compare the simulated and observed data: sum of the difference in number of frogs over all time points:

$$
S_i=\sum_j \sum_k \frac{|x_{ik,obs}(t_j)-x_{i,sim}(t_j|\theta)|}{N_{obs}}
$$

for compartment $x_i$ in observed data $X_{obs}$ (for each experiment $k$) and simulated data $X_{sim}$ for proposed parameter set $\theta$ at time $t_j$ over all observation time points $T = \{1,2,4,6,8,10,15\}$ weeks for $N_{obs}$ observations (7 time points x 4 experiments = 28). We use this statistic to determine if want to accept a proposed parameter set $\theta$ or not.

```{r, metricdef}
# Function to calculate the summary statistic for acceptance
# @param Xobs a dataframe of observed data with a time column named 't' and one column per compartment
# @param Xsim a dataframe of simulated data matching structure of Xobs
# @return S a named vector of summary statistics with one element per compartment
calcSummaryStatisticsOneEnv <- function(Xobs, Xsim) {
  # Check the dataframes are matching structures (colnames and time)
  stopifnot(
    colnames(Xsim) %in% colnames(Xobs), # Order apathetic
    colnames(Xobs) %in% colnames(Xsim), # Order apathetic
    unique(Xobs$t) == unique(Xsim$t)
  )
  # Now we need to calculate the differences for each time point
  # Split the observations into individual time points
  Xdiff <- split(Xobs, Xobs$t) %>%
     lapply(function(xobsi) {
       # Get the relevant simulation row and exclude time column
       xsimi <- Xsim[Xsim$t==xobsi$t[1],names(Xsim) != "t"] 
       # Calculate the difference between obs and sim
       # Note mapply is much faster than data frame subtraction
       # (xobsi[,names(xsimi)] ensures the same column order)
       sdiff <- mapply("-", xobsi[,names(xsimi)], xsimi)
       # Return the absolute values of the differences
       return(abs(sdiff))
     }) %>% 
     # Convert back to a data frame
     do.call(rbind, .)
  # Return the compartment sums / total number of observations
  return(colSums(Xdiff) / nrow(Xobs))
}

calcSummaryStatisticsBothEnv <- function(Xobs, Xsim) {
  # Shaded
  Xobs_sh <- Xobs[Xobs$shaded, colnames(Xobs) != "shaded"]
  Xsim_sh <- Xsim[Xsim$shaded, colnames(Xsim) != "shaded"]
  Si_sh <- calcSummaryStatisticsOneEnv(Xobs_sh, Xsim_sh)
  # Unshaded
  Xobs_un <- Xobs[!Xobs$shaded, colnames(Xobs) != "shaded"]
  Xsim_un <- Xsim[!Xsim$shaded, colnames(Xsim) != "shaded"]
  Si_un <- calcSummaryStatisticsOneEnv(Xobs_un, Xsim_un)
  # Return the average of shaded and unshaded
  return((Si_sh+Si_un)/2)
}
```

### Priors

We also need to choose how we choose our new parameter set. This requires us to set priors on all the parameters. We choose:

-   Transmission rate $\beta$: uniformly distributed \~U(0,1)
-   Vaccination effect $\alpha$: normally distributed around 1 (no effect), truncated at 0, sd = 1? (mainly to make sure we get sufficient sampling at the very low values where we expect it to be).
-   Recovery rate $\omega$: uniformly distributed U(0,7) (max of 1 day?)
-   Death rate $\mu$: Uniformly distributed between 0 and 1, noting the mortality rate across the whole experiment was around 0.02. (Tried using a normal around 0.02 but thought it might be influencing the result too much).

```{r, proposaldef}
# Function to sample a new parameter set
# @param nSamples the number of samples requested
# @return proposed a list of named numeric parameter sets
samplePrior <- function(nSamples)
{
  proposed = lapply(1:nSamples, function(i) {
    c(beta_sh = runif(1, min = 0, max = 1),
      beta_un = runif(1, min = 0, max = 1),
      alpha = rtnorm(1, mean = 1, sd = 1, lower = 0),
      omega = runif(1, min = 0, max = 7),
      mu = runif(1, min = 0, max = 1) )
  })
  return(proposed)
}
```

## Run ABC rejection

Now we run a simple ABC using the prescribed metric and look at a few acceptance functions.

```{r abcparallel}
iterations = 1e5
# First let's generate all the samples and run in parallel
parametersets <- samplePrior(iterations)
# Get the vector of output times from observed data
tOut <- unique(obs$t)
# Run model in parallel for each parameter set
registerDoParallel(4)
run.time <- system.time({
  XsimList <- foreach(p=parametersets) %dopar% {
    runModelBothEnv(params.fit = p, Tvec = tOut)
  }
})
data.frame(as.list(run.time)) %>% 
  kable(caption=paste("Model run computational time for",iterations,"iterations."))
# Calculate the summary statistics
summ.time <- system.time({
SList <- foreach(Xsimi = XsimList) %dopar% {
       calcSummaryStatisticsBothEnv(obs, Xsimi)
}
})
data.frame(as.list(summ.time)) %>% 
  kable(caption=paste("Summary statistic computational time for",iterations,"iterations."))
# Save the results (separately cos XsimList is quite big)
save(parametersets, XsimList, file="data/ABCSimulationData.RData")
save(parametersets, SList, file="data/ABCSummaryStatistic.RData")
```
