---
title: "Run an ABC for the Chytrid Model using synthetic data"
format: 
    html:
        self-contained: true
editor: visual
execute: 
  echo: false
  warning: false
---

```{r, libraries}
#| echo: false
#| warning: false
library(ggplot2)  # For plotting
library(knitr)    # For table output
library(magrittr) 
library(GillespieSSA) # To run gillespie trajectories
library(dplyr)
library(tidyr)
library(msm) # For truncated normal distribution
# Parallel running
library(foreach)
library(doParallel) 

# Parameters that might want to change
iterations = 1e4
set.seed(1) # Set the seed (for gillespie)
compartments = c("S","I1","I2","I3","SV","IV1","IV2","IV3") # Order of compartments for plotting
num_cores = 4
```

## Run Model

We use the Gillespie algorithm to run the model, both for the data generation and the parameter estimation, using the GillespieSSA library.

```{r, modelrundef}
# Source the ODE system from the R script
source("chytrid_models_and_residuals.R")
# Get the gillespie system using the function from this script
gillespie_system <- build_system_gillespie()

# A function that solves the ODE model for a certain set of params
# @param params the parameter guess we want to run
# @param Tvec a vector of times we want to get the output for
# @return data.frame of the compartment counts at requested times
runModel <- function(params.fit, Tvec)
{
  # Initial conditions
  x0 <- c( S=5,  I1=5,  I2=0,  I3=0, 
           SV=5, IV1=5, IV2=0, IV3=0)
  # Combine fitting parameters with other model parameters 
  # (These are defined in chytrid_models_and_residuals.R)
  parameters[names(params.fit)] = params.fit
  parameters[["mu"]] = lsfitparams[["mu"]]
  # Run a trajectory
  trajectory <- ssa(x0 = x0,
      a = gillespie_system[["a"]],
      nu = gillespie_system[["nu"]],
      parms = parameters, 
      tf = max(Tvec),
      method = ssa.d(),verbose = FALSE, consoleInterval = 1)
  # Extract the data at the correct time points
  Xi <- lapply(setNames(Tvec,Tvec),function(ti, data) {
    # Determine the appropriate row in the solution (max is in case ti=0)
    rowi = max(which(data[,"t"] >= ti)[1]-1,1)
    # If it returns NA it will be the last time point
    rowi = ifelse(is.na(rowi), nrow(data), rowi)
    return(data[rowi,])
  }, data=trajectory$data)
  Xi <- bind_rows(Xi, .id="t") # override the t column with obs time
  return(Xi)
}

# A function that solves the ODE model over 2 parameter sets-one per environment (shaded: beta_sh and unshaded: beta_un)
# @param params the parameter guesses we want to run
# @param Tvec a vector of times we want to get the output for
# @return data.frame of the compartment counts at requested times
runModelBothEnv <- function(params.fit, Tvec){
  # Environment independent parameters
  params.env <- params.fit[grep("beta",names(params.fit),invert=T)]
  # Run shaded model
  params.shaded = c(params.env, beta=params.fit[["beta_sh"]])
  Xsim_sh <- runModel(params.shaded, Tvec)
  # Run unshaded model
  params.unshaded = c(params.env, beta=params.fit[["beta_un"]])
  Xsim_un <- runModel(params.unshaded, Tvec)
  # Return both as dataframe with boolean column for shaded or not
  Xsim_sh$shaded = T
  Xsim_un$shaded = F
  return(rbind(Xsim_sh, Xsim_un))
}
```

## Synthetic data generation

We simulate the model, using the parameters from the deterministic fit, using gillespie. We use the same number of observed experiments as the actual experiments (4 shaded, 4 unshaded).

```{r synthetic-data}
# Get the correct parameter values for shaded (_sh) and unshaded (_un)
source("chytrid_models_and_residuals.R")
parameters_ls <- unlist(modifyList(as.list(parameters),as.list(lsfitparams)))
obs.synth <- lapply(1:4,function(i) runModelBothEnv(parameters_ls, Tvec = c(0,1,2,4,6,8,10,15))) %>%
                bind_rows(.id="runid")
obs.synth$t <- as.numeric(obs.synth$t)
pivot_longer(obs.synth, cols = !c(runid,t,shaded)) %>%
mutate(compartment = factor(name,levels=c("S","I1","I2","I3","SV","IV1","IV2","IV3")), shaded = ifelse(shaded,"Shaded","Unshaded")) %>%
    ggplot(aes(x=t,y=value,group=runid,colour=runid)) + geom_line() + geom_point() + facet_grid(shaded~compartment)
```

## Run ABC

### Summary statistic

Now we need to choose the summary statistic(s) we want to use to compare the simulated and observed data: sum of the difference in number of frogs over all time points:

$$
S_i=\sum_j \sum_k \left(x_{ik,obs}(t_j)-x_{i,sim}(t_j|\theta)\right)^2
$$

for compartment $x_i$ in observed data $X_{obs}$ (for each experiment $k$) and simulated data $X_{sim}$ for proposed parameter set $\theta$ at time $t_j$ over all observation time points $T = \{1,2,4,6,8,10,15\}$ weeks for $N_{obs}$ observations (7 time points x 4 experiments = 28). We use this statistic to determine if want to accept a proposed parameter set $\theta$ or not.

```{r, metricdef}
# Function to calculate the summary statistic for acceptance
# @param Xobs a dataframe of observed data with a time column named 't' and one column per compartment
# @param Xsim a dataframe of simulated data matching structure of Xobs
# @return S a named vector of summary statistics with one element per compartment
calcSummaryStatisticsOneEnv <- function(Xobs, Xsim) {
  # Check the dataframes are matching structures (colnames and time)
  stopifnot(
    colnames(Xsim) %in% colnames(Xobs), # Order apathetic
    colnames(Xobs) %in% colnames(Xsim), # Order apathetic
    unique(Xobs$t) == unique(Xsim$t)
  )
  # Now we need to calculate the differences for each time point
  # Split the observations into individual time points
  Xdiff <- split(Xobs, Xobs$t) %>%
     lapply(function(xobsi) {
       # Get the relevant simulation row and exclude time column
       xsimi <- Xsim[Xsim$t==xobsi$t[1],names(Xsim) != "t"] 
       # Calculate the difference between obs and sim
       # Note mapply is much faster than data frame subtraction
       # (xobsi[,names(xsimi)] ensures the same column order)
       sdiff <- mapply("-", xobsi[,names(xsimi)], xsimi)
       # Divide by the average number of frogs at this time point
       avgFrogs <- colMeans(xobsi[names(xsimi)])+1
       sdiff <- apply(sdiff,1,function(si){si/avgFrogs}) %>% t()
       # Return the absolute values of the differences
       return(abs(sdiff))
     }) %>% 
     # Convert back to a data frame
     do.call(rbind, .)
  # Return the sum of the squares of each compartment
  return(colSums(Xdiff^2))
}

# Function to calculate the summary statistic for acceptance for both shaded and unshaded environments
# Essentially calls calcSummaryStatisticsOneEnv defined above twice
# @param Xobs a dataframe of observed data with a time column named 't' and one column per compartment
# @param Xsim a dataframe of simulated data matching structure of Xobs
# @return S a named vector of summary statistics with one element per compartment
calcSummaryStatisticsBothEnv <- function(Xobs, Xsim) {
  # Shaded
  Xobs_sh <- Xobs[Xobs$shaded, colnames(Xobs) != "shaded"]
  Xsim_sh <- Xsim[Xsim$shaded, colnames(Xsim) != "shaded"]
  Si_sh <- calcSummaryStatisticsOneEnv(Xobs_sh, Xsim_sh)
  # Unshaded
  Xobs_un <- Xobs[!Xobs$shaded, colnames(Xobs) != "shaded"]
  Xsim_un <- Xsim[!Xsim$shaded, colnames(Xsim) != "shaded"]
  Si_un <- calcSummaryStatisticsOneEnv(Xobs_un, Xsim_un)
  # Return the sum of shaded and unshaded
  return(Si_sh+Si_un)
}
```

### Priors

We also need to choose how we choose our new parameter set. This requires us to set priors on all the parameters. We choose:

-   Transmission rate $\beta$: uniformly distributed \~U(0,1)
-   Vaccination effect $\alpha$: normally distributed around 1 (no effect), truncated at 0, sd = 1? (mainly to make sure we get sufficient sampling at the very low values where we expect it to be).
-   Recovery rate $\omega$: uniformly distributed U(0,7) (max of 1 day?)
-   Death rate $\mu$: Uniformly distributed between 0 and 1, noting the mortality rate across the whole experiment was around 0.02. (Tried using a normal around 0.02 but thought it might be influencing the result too much).

```{r, proposaldef}
# Function to sample a new parameter set
# @param nSamples the number of samples requested
# @return proposed a list of named numeric parameter sets
samplePrior <- function(nSamples)
{
  proposed = lapply(1:nSamples, function(i) {
    c(beta_sh = runif(1,min=0.05*0.9, max=0.05*2), #runif(1, min = 0, max = 1),
      beta_un = runif(1,min=0.02*0.9,max=0.02*2), #runif(1, min = 0, max = 1),
      alpha = runif(1,min=0.177*0.9,max=0.177*2), #rtnorm(1, mean = 1, sd = 1, lower = 0),
      omega = 0.258, #runif(1,min=0.258*0.9,max=0.258*2),#runif(1, min = 0, max = 7),
      #mu = runif(1, min = 0, max = 1),
      iteration = i)
  })
  return(proposed)
}
```

## Run ABC rejection

Now we run a simple ABC using the prescribed metric and look at a few acceptance functions.

```{r abcparallel}
# First let's generate all the samples and run in parallel
parametersets <- samplePrior(iterations)
# Get the vector of output times from observed data
tOut <- unique(obs.synth$t)
# Run model in parallel for each parameter set
registerDoParallel(num_cores)
run.time <- system.time({
  XsimList <- foreach(p=parametersets) %dopar% {
     Xsim = runModelBothEnv(params.fit = p, Tvec = tOut)
     Xsim$iteration = p[["iteration"]]
     return(Xsim)
  }
})
data.frame(as.list(run.time)) %>% 
  kable(caption=paste("Model run computational time for",iterations,"iterations."))
# Calculate the summary statistics
summ.time <- system.time({
SList <- foreach(Xsimi = XsimList) %dopar% {
            obs <- dplyr::select(obs.synth,-runid)
            sim <- dplyr::select(Xsimi,-iteration)
            c(iteration = Xsimi$iteration[1],
              calcSummaryStatisticsBothEnv(obs,sim))
        }
})
Si <- do.call(rbind, SList) %>% data.frame()
data.frame(as.list(summ.time)) %>% 
  kable(caption=paste("Summary statistic computational time for",iterations,"iterations."))
```

## Analysis

### Acceptance

Let us accept the top 2.5% of the simulations, using the maximum error across the compartments and the sum of the error over all simulations.

```{r acceptanceplots}
N_accepted = round(2.5*iterations/100)
it_colid = which(names(Si)=="iteration")
sum_error <- apply(Si[,-it_colid], 1, sum)
max_error <- apply(Si[,-it_colid], 1, max)

accepted.sum <- parametersets[order(sum_error)[1:N_accepted]]
accepted.max <- parametersets[order(max_error)[1:N_accepted]]

accepted.sum <- do.call(rbind,accepted.sum)
accepted.max <- do.call(rbind,accepted.max)

actualp <- data.frame(name=names(lsfitparams),v=as.numeric(lsfitparams))

pivot_longer(data.frame(accepted.sum),cols=!iteration) %>%
    ggplot(aes(x=value)) + geom_histogram() + facet_wrap(~name, scales="free") +
    geom_vline(aes(xintercept=v), colour="#ff5321", data=actualp) +
    labs(title="Sum of squares error")

# pivot_longer(data.frame(accepted.max),cols=!iteration) %>%
#     ggplot(aes(x=value)) + geom_histogram() + facet_wrap(~name, scales="free")+
#     geom_vline(aes(xintercept=v), colour="#ff5321", data=actualp) +
#     labs(title="Maximum square error across the different compartments")
```

```{r}
dplyr::summarise(as.data.frame(accepted.sum),across(!iteration,~median(.x))) %>% knitr::kable(caption="median")

dplyr::summarise(as.data.frame(accepted.sum),across(!iteration,~mean(.x))) %>% knitr::kable(caption="mean")
```

Error profile (squared error) over all simulated data, fill indicates error in accepted parameter sets.

```{r, errorhistogram}
#| fig.width = 20
mutate(Si,
      accepted = iteration %in% accepted.sum[,"iteration"]) %>%
    pivot_longer(cols=!c(iteration,accepted)) %>%
    mutate(name = factor(name, levels=compartments)) %>%
      ggplot() + 
      geom_histogram(aes(x=value,y=0.5*after_stat(density),fill=accepted),position=position_dodge()) +
      facet_wrap(~name, scales="free", ncol=4) + 
scale_fill_manual(values=c("darkblue","seagreen")) + labs(title="Condition: Sum of errors")


mutate(Si,
      accepted = iteration %in% accepted.max[,"iteration"]) %>%
    pivot_longer(cols=!c(iteration,accepted)) %>%
    mutate(name = factor(name, levels=compartments)) %>%
      ggplot() + 
     geom_histogram(aes(x=value,y=0.5*after_stat(density),fill=accepted),position=position_dodge()) +
      facet_wrap(~name, scales="free", ncol=4) + 
scale_fill_manual(values=c("darkblue","seagreen")) + labs(title="Condition: Maximum error")
```

### Spaghetti plots

```{r spaghettiplots}
N_realisations = N_accepted
# Convert data to long form and set correct compartment order
r <- XsimList[order(sum_error)[1:N_realisations]] %>% 
        do.call(rbind,.) %>%
        pivot_longer(cols=!c(t,shaded,iteration)) %>%
        mutate(name = factor(name,levels=compartments), t = as.numeric(t))
obs.synth.long <- pivot_longer(obs.synth, cols=!c(runid, t, shaded)) %>%
        mutate(name = factor(name,levels=compartments))
# Plot
ggplot(r,aes(x=t,y=value)) +
    geom_line(aes(group=iteration,colour="simulated"), alpha=0.1)  +
    geom_line(aes(group=runid, colour="observed"), data=obs.synth.long) + 
    facet_grid(shaded~name)

```

```{r acceptanceplots2}
N_accepted = round(1*iterations/100)
it_colid = which(names(Si)=="iteration")
sum_error <- apply(Si[,-it_colid], 1, sum)
max_error <- apply(Si[,-it_colid], 1, max)

accepted.sum <- parametersets[order(sum_error)[1:N_accepted]]
accepted.max <- parametersets[order(max_error)[1:N_accepted]]

accepted.sum <- do.call(rbind,accepted.sum)
accepted.max <- do.call(rbind,accepted.max)

actualp <- data.frame(name=names(lsfitparams),v=as.numeric(lsfitparams))

pivot_longer(data.frame(accepted.sum),cols=!iteration) %>%
    ggplot(aes(x=value)) + geom_histogram() + facet_wrap(~name, scales="free") +
    geom_vline(aes(xintercept=v), colour="#ff5321", data=actualp) +
    labs(title="Sum of squares error")
```
